Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
move code around, another fix

class BoxedFloat(Base):
    def __init__(self, floatval):
        self.floatval = floatval
    def add(self, other):
        return other.add__float(self.floatval)
    def add__int(self, intother):
        floatvalue = float(intother) + self.floatval
        return BoxedFloat(floatvalue)
    def add__float(self, floatother):
        return BoxedFloat(floatother + self.floatval)
    def is_positive(self):
        return self.floatval > 0.0
\end{verbatim}
\caption{A simple object model}
\label{fig:objmodel}
\end{figure}

@@

To understand the problems more directly, let us consider a simple function
that uses the object model:

@@

\begin{verbatim}
def f(y):
    res = BoxedInteger(0)
    while y.is_positive():
        res = res.add(y).add(BoxedInteger(-100))
        y = y.add(BoxedInteger(-1))
    return res
\end{verbatim}

The loop iterates \texttt{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \texttt{is\_positive} and even two for each
call to \texttt{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \texttt{f} is simply a sequence of
float or integer additions.

\begin{figure}
\texttt{
\begin{tabular}{l}
\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res \\
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/get} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

>>>>

class BoxedFloat(Base):
    def __init__(self, floatval):
        self.floatval = floatval
    def add(self, other):
        return other.add__float(self.floatval)
    def add__int(self, intother):
        floatvalue = float(intother) + self.floatval
        return BoxedFloat(floatvalue)
    def add__float(self, floatother):
        return BoxedFloat(floatother + self.floatval)
    def is_positive(self):
        return self.floatval > 0.0

def f(y):
    res = BoxedInteger(0)
    while y.is_positive():
        res = res.add(y).add(BoxedInteger(-100))
        y = y.add(BoxedInteger(-1))
    return res
\end{verbatim}
\caption{A Simple Object Model and an Example Function Using it}
\label{fig:objmodel}
\end{figure}

@@

To understand the problems more directly, let us consider the simple function
\texttt{f} that uses the object model (see the bottom of
Figure~\ref{fig:objmodel}).

@@

The loop in \texttt{f} iterates \texttt{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \texttt{is\_positive} and even two for each
call to \texttt{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \texttt{f} is simply a sequence of
float or integer additions.

\begin{figure}
\texttt{
\begin{tabular}{l}
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \texttt{y} and \texttt{res} in the original function.

The trace shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/get} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.
